---
title: "RwithSQL"
author: "Marco Langenhuizen"
date: "2018 M08 29"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
con <- DBI::dbConnect(
  odbc::odbc(),
  Driver    = "SQL Server",
  Server    = "mssql.fhict.local",
  Database  = "dbi884568_klantartik",
  UID       = "dbi884568_klantartik",
  PWD       = "kl@nt4artikel", #rstudioapi::askForPassword("Database password"),
  Port      = 1433
)
```

```{sql, connection=con}
select woonplaats, year(besteldatum) as jaar, month(besteldatum) as maand,
  count(*) as aantal
from klant k
join bestelling b
  on k.klantnr = b.klantnr
group by woonplaats, year(besteldatum), month(besteldatum)
order by woonplaats, jaar, maand
;
```

We can also attach the resultset of an SQL query to an R variable. The resultset is returned as a dataframe.

```{sql, connection=con, output.var=df_sales}
select b.bestelnr, klantnr, verkoper, besteldatum, leverdatum, betaaldatum, aantal, r.artikelnr, a.catcode, prijs * aantal as prijs
from bestelling b
join bestelregel r
  on b.bestelnr = r.bestelnr
join artikel a
  on r.artikelnr = a.artikelnr
join artikelcategorie c
  on a.catcode = c.catcode
join artikelprijs p
  on r.artikelnr = p.artikelnr
  and besteldatum between begindatum and einddatum
;
```

Now print the query result.

```{r}
df_sales
```

In the result we can see that all dates have a character string datatype. We want them to be proper dates so that we can use all date related functions such as `month()` and `year()`. We change the types using a `dplyr::mutate()`.

```{r}
library(lubridate)

df_sales <- df_sales %>% 
  mutate(
    besteldatum=as_date(besteldatum),
    leverdatum=as_date(leverdatum),
    betaaldatum=as_date(betaaldatum)
  )
```

Now we should be able to use lubridate's diverse set of date functions.

```{r}
df_sales %>% 
  group_by(jaar=year(besteldatum),  maand=month(besteldatum, label=TRUE)) %>% 
  summarise(totaal=sum(prijs))
```

Now that our dates are proper R dates, we can use lubridate's date/time functios.

```{r}
df_sales %>% 
  filter(year(besteldatum) == 2014) %>% 
  group_by(maand=month(besteldatum, label=TRUE)) %>% 
  summarise(totaal=sum(prijs))
```

Note that the result type of `month(besteldatum, label=TRUE)` is ordered factor (ord). If we would make a plot with month on axis, the order will be ok.

```{r}
library(dbplot)

df_sales %>% 
  filter(year(besteldatum) == 2014) %>% 
  group_by(maand=month(besteldatum, label=TRUE)) %>% 
  summarise(totaal=sum(prijs)) %>% 
  dbplot_bar(maand, totaal)
```

The dbplot package is a layer over the ggplot plotting package. Dbplot offers us a small subset of ggplot's plotting functionality by pushing the calculation of the plot dataset to the database back-end. In the query above we were explicit about most of the back-end calculation, but dbplot can figure out a lot by itself. Alternatively to the query above, we could have specified:

```{r}
df_sales %>% 
  filter(year(besteldatum) == 2014) %>% 
  dbplot_bar(month(besteldatum, label=TRUE), sum(prijs))
```

Closing RStudio automatically disconnects.

```{r}
#DBI::dbDisconnect(con)
```

